### 2023-05-08 chapter0/chapter0.tex
Added citation for hemoglobin background

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary ``receptor'' and ``donor'' sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in part a result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

@@

In some cases, folded domains have multiple structures, or conformations, which are related to different functional states. For example, hemoglobin has two forms, traditionally called the T and R states. The T state is hemoglobin's oxygen-free form, but on binding oxygen its structure shifts to the R state. The change is small, differing at most by only a few hydrogen atoms. However, this enough to re-orient the atoms that interact with oxygen, allowing it to bind more tightly and promoting oxygen uptake at the other three binding sites. Once the red blood cells reach their destination, other physiological factors favor the adoption of the T state, which coordinates the release of all four oxygen atoms. In the other cases, conformational changes can dramatically re-organize a protein's structure. For example, the 26S proteasome is a complex of proteins responsible for degrading other proteins. Its structure is highly complex and consists of over three dozen distinct protein subunits, which are in turn organized into three subcomplexes: a lid, a base, and a core~\cite{Finley2016, Bard2018}. The functions of these subcomplexes are roughly analogous to the parts of a paper shredder. The lid is like the outer shell because it regulates access to the ``motor'' in the base and the ``blades'' in the core that pull in and degrade the protein, respectively.\footnote{As with many analogies, this comparison to a paper shredder simplifies several structural and functional details of the proteasome. For example, in a paper shredder the motor powers the blades which both pull in and shred the paper. In the proteasome, however, these are distinct steps. The motor subunits in the base first physically interact with the target protein to simultaneously unfold and pull it into the core. Different subunits in the core then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid. Unlike a paper shredder, however, the motor is activated by a conformational change rather than a physical switch. When the tail of a protein marked for degradation is inserted in the motor, the lid shifts by nearly forty hydrogen atoms to align the motor with the pore that leads into the core subcomplex. Despite the scale of this re-arrangement, it occurs over the span of only half a second~\cite{Bard2019}. Thus, when folded domains have multiple conformations, they are generally discrete forms without stable intermediates.

>>>>

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids.\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the 20 ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.} Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary ``receptor'' and ``donor'' sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in part a result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for valine~\cite{Voet2015}. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

@@

In some cases, folded domains have multiple structures, or conformations, which are related to different functional states. For example, hemoglobin has two forms, traditionally called the T and R states~\cite{Voet2015}. The T state is hemoglobin's oxygen-free form, but on binding oxygen its structure shifts to the R state. The change is small, differing at most by only a few hydrogen atoms. However, this enough to re-orient the atoms that interact with oxygen, allowing it to bind more tightly and promoting oxygen uptake at the other three binding sites. Once the red blood cells reach their destination, other physiological factors favor the adoption of the T state, which coordinates the release of all four oxygen atoms. In the other cases, conformational changes can dramatically re-organize a protein's structure. For example, the 26S proteasome is a complex of proteins responsible for degrading other proteins. Its structure is highly complex and consists of over three dozen distinct protein subunits, which are in turn organized into three subcomplexes: a lid, a base, and a core~\cite{Finley2016, Bard2018}. The functions of these subcomplexes are roughly analogous to the parts of a paper shredder. The lid is like the outer shell because it regulates access to the ``motor'' in the base and the ``blades'' in the core that pull in and degrade the protein, respectively.\footnote{As with many analogies, this comparison to a paper shredder simplifies several structural and functional details of the proteasome. For example, in a paper shredder the motor powers the blades which both pull in and shred the paper. In the proteasome, however, these are distinct steps. The motor subunits in the base first physically interact with the target protein to simultaneously unfold and pull it into the core. Different subunits in the core then break the exposed chemical bonds between amino acid residues in the chain.} Like a paper shredder, the motor is only engaged when a protein is correctly positioned in the lid. Unlike a paper shredder, however, the motor is activated by a conformational change rather than a physical switch. When the tail of a protein marked for degradation is inserted in the motor, the lid shifts by nearly forty hydrogen atoms to align the motor with the pore that leads into the core subcomplex. Despite the scale of this re-arrangement, it occurs over the span of only half a second~\cite{Bard2019}. Thus, when folded domains have multiple conformations, they are generally discrete forms without stable intermediates.
